A Computational Account of Default Mode Function\\ by Reinforcement Learning Theory

### 2017-03-26 paper/paper.tex
more

\section{Known Neurobiological Properties of the Default Mode Network}
\textbf{XXX This section is presented as transfinite block of concatenated text.
  Please reorganize into subsections and paragraphs!}
We will first attempt to deconstruct the overall DMN function
by the mechanistic relevance of each individual network node
based on published findings from neuroscience experiments.
%
\subsection{The PMC and PFC}
The posteriormedial cortex (PMC) and
the dorsomedial prefrontal cortex (dmPFC) within the DMN
are responsible for the highest metabolic turn-over
of glucose energy consumption in humans \citep{raichle2001pnas}.
The PMC myelinates relatively late during postnatal development in monkeys
\citep{goldman1987development}, generally considered to
be a sign of phylogenetic sophistication (Flechsig, 1920).
Physiological and disturbed metabolic fluctuations in the
human PMC have been repeatedly related to
phenomena of changed conscious awareness,
including anesthesia,
sleep, and forms of coma \citep{cavanna2006precuneus}.
%
The PMC has long been speculated to reflect constant computation of
the environmental statistics and its internal representation
as an "inner minds eye" \citep{cavanna2006precuneus}.
For instance, B\'alint's syndrome is a neurological disorder of conscious
awareness that results from damage in the bilateral parietal cortex
\citep{balint1909seelenlahmung}.
Patients are plagued by an
inability to bind various individual features of the visual
environment into an integrated whole (i.e., simultanagnosia)
as well as inability to direct action towards
currently unattended environmental objects
(i.e., optic ataxia).
This can be viewed as a high-level impairment in the gathering
of information about alternative objects (i.e., exploration) as well as
leveraging these environmental opportunities (i.e., exploitation).
Congruently,
the human PMC was coupled in two functional connectivity modalities
with the amygdala
(involved in significance evaluation) and
the nucleus accumbens (involved in reward evaluation).
Specifically, the human ventral posterior cingulate cortex was
the PMC region most connected to the laterobasal
(rather than centromedial or superficial) nuclei group
of the amygdala \citep{bzdok2015subspecialization}.
\textbf{XXX: the previous sentence seems to be bad / incomplete grammar!}
This amygdalar subregion has been proposed to
continuously scan environmental input
for biological significance assessment.
%
Indeed,
electrophysiological recordings in animals implicated the PMC in
strategic selection \citep{pearson2009neurons},
risk assessment \citep{mccoy2005risk},
and outcome-contingent behavioral modulation \citep{hayden2009electrophysiological},
while its retrosplenial portion was
more specifically implicated in approach-avoidance behavior
\citep{vann2009does}.
Neuron spiking activity in the PMC allowed distinguishing
whether a monkey would persue an exploratory or exploitative
behavioral strategy in a complex food foraging task \citep{pearson2009neurons}.
Single-cell recordings in the monkey PMC
demonstrated this brain region's sensitivity to
subjective target utility \citep{mccoy2005risk} and integration
across individual decision-making instances \citep{pearson2009neurons}.

@@

The HC is well known to be involved in memory and
spatial navigation in animals and humans.
Its highly recursive anatomical architecture
may be specifically designed to allow reconstructing
entire episodes of experience from memory fragments.
%
While the HC in the medial temporal lobe system
is traditionally believed to allow remembering the past,
there is now increasing evidence for a role
in constructing mental models in general
\citep{maguire2016, schacter2007remembering}.
Indeed,
hippocampal damage is
not only associated with an impairment in reexperiencing the past (i.e., amnesia),
but also thinking about one’s own future and
imagining new fictitious experiences more broadly \citep{hassabis2007patients}.
Mental scenes created by hippocampus-lesioned patients exposed a lack of
spatial integrity, richness in detail, and overall coherence.
%
Single-cell recordings in the animal hippocampus revealed
some constantly active neuronal ensembles whose firing coincided with
distinct locations in space while the animal navigated through its surroundings.
London taxi drivers, individuals with high performance in spatial navigation,
were shown to exhibit increased grey matter volume in the
posterior hippocampus \citep{maguire2000navigation}.
But encoding and generative reconstruction in the hippocampus extends
beyond mere spatial knowledge of the environment.
Based on large-scale recordings of hippocampal neuronal populations,
complex spiking patterns can be followed across extended periods including
their modification of input-free self-generated patterns
after environmental events \citep{buzsaki2004large}.
Specific spiking sequences, that were elicited in experimental task conditions,
have been shown to be reenacted spontaneously during
quiet wakefulness and sleep \citep{hartley2014space}.
Moreover, spike sequences measured in hippocampal place cells of rats
featured reoccurrence directly after experimental trials
as well as directly before upcoming experimental trials \citep{diba2007forward}.
Such hippocampal ensemble burts during rest and sleep
have been proposed to be critical in communicating local information
to the neocortex for long-term storage, potentially also in the nodes of the DMN.
These mechanisms of the hippocampus probably make important contributions to the
recollection of autobiographical memory episodes and other
reexperienced or newly generated mental scenarios
\citep{hassabis2007patients}.
%
The HC thus orchestrates elements of experienced environmental aspects for
consolidations based on reenactment and for integration into
rich mental scene construction. In this view, the HC my even influence
ongoing perception in the current environment
\citep{maguire2016}.
% (TODO: relate to self-play in deepmind Go).
% (TODO: Kai had an interesting idea: the deja-vu effect
% could be when you already accidentally perturbed past
% such that it really occures in reality later.)

\subsection{The TPJ: Situation understanding and context-switching}
Finally,
the right and left TPJ are known to have hemispheric differences
according to their cytoarchitectonic borders and gyrification pattern
\citep{seghier2013angular}.
Neuroscientific investigations on hemispheric functional specialization
have converged to the right versus left cerebral hemisphere as dominant for
attention versus language functions.
%
The aRTPJ (\textbf{XXX ???})  is a key node network is central for
action initiation during externally structured tasks and
sensorimotor control by integrating supramodal stimulus-guided attention
\citep{corbetta2002control}. \textbf{XXX Previous sentence is weird!}
Involvement of this DMN node was repeatedly reported in
multi-step action execution \citep{hartmann2005takes},
visuo-proprioceptive conflict (Balslev et al., 2005) and
multi-modal detection of sensory changes across
visual, auditory, or tactile stimulation in a multi-modal fMRI study
\citep{downar2000multimodal}.
In humans, direct electrical stimulation of the
RTPJ during neurosurgery was associated with altered perception
and stimulus awareness \citep{blanke2002neuropsychology}.
%
Importantly, the RTPJ has been shown to be intimately related to
supramodal prediction and error signaling.
It was argued that the RTPJ encodes actions and ensuing outcomes
without necessarily relating those to outcome value
\citep{liljeholm2013neural, hamilton2008action,
jakobs2009effects}.
Neural activity in the RTPJ has been argued to be responsible
for stimulus-driven attentional reallocation to
salient and surprising sources of information
as a ‘‘circuit breaker’’ that recalibrates control and maintenance systems
\citep{bzdok2013tpj, corbettashul2008}.
% (Bzdok et al., 2013; Corbetta et al., 2008).
Indeed, patients with right TPJ damage have particular difficulties
with multistep actions \citep{hartmann2005takes}.
In the face of large discrepancies between actual and previously predicted
environmental events the RTPJ acts a potential switch between
externally-oriented mind sets focussed on the
sensory world and internally-oriented mind sets focussed
on self-relevant mind-wandering of possible mental scenarios.
Transient RTPJ in humans for instance disrupted the
impact of predicted intentions of other individuals
\citep{young2010disruption},
a capacity believed to be subserved by the DMN.
The RTPJ might hence be an important player that shifts away
from the default active ‘‘internally directed’’ processes
to deal instead with immediate environmental objects and contexts.
% (TODO: slow endogeneous changes in state transitions by the policy and value matrices)

@@

@@

\section{Reinforcement Learning as a computational model for DMN function}
The main idea of this section is to argue that the DMN has biological substrates (gradient
computation, backprop, experience replay, etc.) for all the
major components of an efficient \textit{Reinforcement Learning} algorithm, and all these
components integrate in a plausible way.
Reinforcement Learning (RL) is learning to achieve a goal by interacting with an external
environment. In a general RL framework, an \textit{agent}
interacts with an  environment in a trial-by-error fashion by taking
actions / decisions $a$, which trigger changes in the
\textit{state} of the environment
$s \rightarrow s'$, accompanied by a real-valued \textit{reward}
$r = r(s, a)$ (or regret!, since it can be negative) collected by the
agent. In this view, the environmental is partially controlled by
the action of the agent and the reward should be thought
of as an instantaneous satisfaction accompanying the execution of
an action, and may be delayed, only becoming substantial when accumulated over an extended period of time.
%% Here the subscript $t$ denotes time, and can be discrete or
%% continuous.
The environment is generally taken as stochastic,
is partially observed, in the sense that only part of the current state is observed by
the agent. This unpredictability
% (in a first-order sense)
is realistic in a model which sets out
to explain DMN function. Indeed, it is known that, the more the external world is predictable,
the more mental activity becomes detached from the actual sensorium
\citep{antrobus1966studies, pope1978regulation}.
Conversely, the more the currently executed task is unknown and unpracticed,
the lower the DMN activity, i.e the tendency for stimulus-independent thoughts
\citep{filler1973daydreaming, teasdale1995stimulus}.

\subsection{Value functions and policies}
The behavior of the agent is governed by some kind of \textit{policy}, which maps states to
sampling probabilities over candidate actions to perform from this state. Starting a time $t=0$,
a policy $\pi$ generates a trajectory as follows
\begin{eqnarray*}
  \begin{split}
    S_1 &\sim p(.|S_0,A_0), \text{ collect rewards }R_0 = r(S_0, A_0)\\
    A_1 &\sim \pi(.|S_1)\\
    S_2 &\sim p(.|S_1,A_1), \text{ collect rewards }R_1 = r(S_1, A_1)\\
    \ldots\\
    A_{t} &\sim \pi(.|S_{t})\\
    S_{t+1} &\sim p(.|S_{t},A_{t}), \text{ collect rewards }R_{t} = r(S_{t}, A_{t})\\
    \ldots
  \end{split}
\end{eqnarray*}
Since an action taken now might have repercussions long into the feature, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
cumulative reward estimate which takes into account the future. A standard for
modeling this accumulation is the so-called
time-discounted \textit{cumulative reward} function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t.
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic).}  measures the cumulative reward of
following a
policy $\pi$.
\begin{mdframed}
  \textbf{Where is ``value'' buffered in the brain ?}
  The amygdala is known to be involved in significance evaluation) and
  the nucleus accumbens is involved in reward evaluation.
  \end{mdframed}
The  goal of RL is then adapt this policy so as to  maximize the cumulative rewards.
In the definition of $G^\pi$ above, the constant $\gamma \in [0, 1]$ is the \textit{discount factor}. The choice $\gamma = 0$   corresponds to a perfectly myopic agent who is solely concerned about their immediate rewards and has no horizon
for the future. For there to be any learning at all, a technical requirement is that $\gamma < 1$.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
Thus the agent is positively biased towards considering rewards which are will
come soon. The reward importance hence decays exponentially
with distance in the future,
given that there is more uncertainty in the farsighted future.
The reward discount factor $\gamma$ can thus be seen as calibrating
the delay discounting behavior of the intelligent agent.
It is also important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
For instance, a deterministic strategy in playing rock-paper-scissors can
be easily exploited and a uniformly random choice is optimal.

>>>>

\section{Known Neurobiological Properties of the Default Mode Network}
\textbf{XXX This section is presented as transfinite block of concatenated text.
  Please reorganize into subsections and paragraphs!}
We will first attempt to deconstruct the overall DMN function
by the mechanistic relevance of each individual network node
based on published findings from neuroscience experiments.
%
\subsection{The posteriormedial cortex (PMC)}
The posteriormedial cortex (PMC) and
the dorsomedial prefrontal cortex (dmPFC) within the DMN
are responsible for the highest metabolic turn-over
of glucose energy consumption in humans \citep{raichle2001pnas}.
The PMC myelinates relatively late during postnatal development in monkeys
\citep{goldman1987development}, generally considered to
be a sign of phylogenetic sophistication (Flechsig, 1920).
Physiological and disturbed metabolic fluctuations in the
human PMC have been repeatedly related to
phenomena of changed conscious awareness,
including anesthesia,
sleep, and forms of coma \citep{cavanna2006precuneus}.
%
The PMC has long been speculated to reflect constant computation of
the environmental statistics and its internal representation
as an "inner minds eye" \citep{cavanna2006precuneus}.
For instance, B\'alint's syndrome is a neurological disorder of conscious
awareness that results from damage in the bilateral parietal cortex
\citep{balint1909seelenlahmung}.
Patients are plagued by an
inability to bind various individual features of the visual
environment into an integrated whole (i.e., simultanagnosia)
as well as inability to direct action towards
currently unattended environmental objects
(i.e., optic ataxia).
This can be viewed as a high-level impairment in the gathering
of information about alternative objects (i.e., exploration) as well as
leveraging these environmental opportunities (i.e., exploitation).
Congruently,
the human PMC was coupled in two functional connectivity modalities
with the amygdala
(involved in significance evaluation) and
the nucleus accumbens (involved in reward evaluation).
Specifically, the human ventral posterior cingulate cortex was
the PMC region most connected to the laterobasal
(rather than centromedial or superficial) nuclei group
of the amygdala \citep{bzdok2015subspecialization}.
\textbf{XXX: the previous sentence seems to be bad / incomplete grammar!}
This amygdalar subregion has been proposed to
continuously scan environmental input
for biological significance assessment.
%
Indeed,
electrophysiological recordings in animals implicated the PMC in
strategic selection \citep{pearson2009neurons},
risk assessment \citep{mccoy2005risk},
and outcome-contingent behavioral modulation \citep{hayden2009electrophysiological},
while its retrosplenial portion was
more specifically implicated in approach-avoidance behavior
\citep{vann2009does}.
Neuron spiking activity in the PMC allowed distinguishing
whether a monkey would persue an exploratory or exploitative
behavioral strategy in a complex food foraging task \citep{pearson2009neurons}.
Single-cell recordings in the monkey PMC
demonstrated this brain region's sensitivity to
subjective target utility \citep{mccoy2005risk} and integration
across individual decision-making instances \citep{pearson2009neurons}.

@@

The HC (\textbf{``HC'' = ?}) is well known to be involved in memory and
spatial navigation in animals and humans.
Its highly recursive anatomical architecture
may be specifically designed to allow reconstructing
entire episodes of experience from memory fragments.
%
While the HC in the medial temporal lobe system
is traditionally believed to allow remembering the past,
there is now increasing evidence for a role
in constructing mental models in general
\citep{maguire2016, schacter2007remembering}.
Indeed,
hippocampal damage is
not only associated with an impairment in reexperiencing the past (i.e., amnesia),
but also thinking about one’s own future and
imagining new fictitious experiences more broadly \citep{hassabis2007patients}.
Mental scenes created by hippocampus-lesioned patients exposed a lack of
spatial integrity, richness in detail, and overall coherence.
%
Single-cell recordings in the animal hippocampus revealed
some constantly active neuronal ensembles whose firing coincided with
distinct locations in space while the animal navigated through its surroundings.
London taxi drivers, individuals with high performance in spatial navigation,
were shown to exhibit increased grey matter volume in the
posterior hippocampus \citep{maguire2000navigation}.
But encoding and generative reconstruction in the hippocampus extends
beyond mere spatial knowledge of the environment.
Based on large-scale recordings of hippocampal neuronal populations,
complex spiking patterns can be followed across extended periods including
their modification of input-free self-generated patterns
after environmental events \citep{buzsaki2004large}.
Specific spiking sequences, that were elicited in experimental task conditions,
have been shown to be reenacted spontaneously during
quiet wakefulness and sleep \citep{hartley2014space}.
Moreover, spike sequences measured in hippocampal place cells of rats
featured reoccurrence directly after experimental trials
as well as directly before upcoming experimental trials \citep{diba2007forward}.
Such hippocampal ensemble burts during rest and sleep
have been proposed to be critical in communicating local information
to the neocortex for long-term storage, potentially also in the nodes of the DMN.
These mechanisms of the hippocampus probably make important contributions to the
recollection of autobiographical memory episodes and other
reexperienced or newly generated mental scenarios
\citep{hassabis2007patients}.
%
The HC thus orchestrates elements of experienced environmental aspects for
consolidations based on reenactment and for integration into
rich mental scene construction. In this view, the HC my even influence
ongoing perception in the current environment
\citep{maguire2016}.
% (TODO: relate to self-play in deepmind Go).
% (TODO: Kai had an interesting idea: the deja-vu effect
% could be when you already accidentally perturbed past
% such that it really occures in reality later.)

\subsection{The TPJ: Situation-understanding and context-switching}
Finally,
the right and left TPJ are known to have hemispheric differences
according to their cytoarchitectonic borders and gyrification pattern
\citep{seghier2013angular}.
Neuroscientific investigations on hemispheric functional specialization
have converged to the right versus left cerebral hemisphere as dominant for
attention versus language functions.
%
The aRTPJ (\textbf{XXX ???})  is a key node network is central for
action initiation during externally structured tasks and
sensorimotor control by integrating supramodal stimulus-guided attention
\citep{corbetta2002control}. \textbf{XXX Previous sentence is weird!}
Involvement of this DMN node was repeatedly reported in
multi-step action execution \citep{hartmann2005takes},
visuo-proprioceptive conflict (Balslev et al., 2005) and
multi-modal detection of sensory changes across
visual, auditory, or tactile stimulation in a multi-modal fMRI study
\citep{downar2000multimodal}.
In humans, direct electrical stimulation of the
RTPJ during neurosurgery was associated with altered perception
and stimulus awareness \citep{blanke2002neuropsychology}.
%
Importantly, the RTPJ has been shown to be intimately related to
supramodal prediction and error signaling.
It was argued that the RTPJ encodes actions and ensuing outcomes
without necessarily relating those to outcome value
\citep{liljeholm2013neural, hamilton2008action,
jakobs2009effects}.
Neural activity in the RTPJ has been argued to be responsible
for stimulus-driven attentional reallocation to
salient and surprising sources of information
as a ‘‘circuit breaker’’ that recalibrates control and maintenance systems
\citep{bzdok2013tpj, corbettashul2008}.
% (Bzdok et al., 2013; Corbetta et al., 2008).
Indeed, patients with right TPJ damage have particular difficulties
with multistep actions \citep{hartmann2005takes}.
In the face of large discrepancies between actual and previously predicted
environmental events the RTPJ acts a potential switch between
externally-oriented mind sets focussed on the
sensory world and internally-oriented mind sets focussed
on self-relevant mind-wandering of possible mental scenarios.
Transient RTPJ in humans for instance disrupted the
impact of predicted intentions of other individuals
\citep{young2010disruption},
a capacity believed to be subserved by the DMN.
The RTPJ might hence be an important player that shifts away
from the default active ‘‘internally directed’’ processes
to deal instead with immediate environmental objects and contexts.
% (TODO: slow endogeneous changes in state transitions by the policy and value matrices)

@@

\section{Reinforcement Learning (RL) as a computational model for DMN function}
We now argue that the DMN has biological substrates (gradient
computation, backprop, experience replay, etc.) for all the
major components of an efficient \textit{Reinforcement Learning} algorithm, and all these
components integrate in a plausible way.
Reinforcement Learning (RL) is learning to achieve a goal by interacting with an external
environment. In a general RL framework, an \textit{agent}
interacts with an  environment in a trial-by-error fashion by taking
actions / decisions $a$, which trigger changes in the
\textit{state} of the environment
$s \rightarrow s'$, accompanied by a real-valued \textit{reward}
$r = r(s, a)$ (or regret!, since it can be negative) collected by the
agent. In this view, the environmental is partially controlled by
the action of the agent and the reward should be thought
of as an instantaneous satisfaction accompanying the execution of
an action, and may be delayed, only becoming substantial when accumulated over an extended period of time.
%% Here the subscript $t$ denotes time, and can be discrete or
%% continuous.
The environment is generally taken as stochastic,
is partially observed, in the sense that only part of the current state is observed by
the agent. This unpredictability
% (in a first-order sense)
is realistic in a model which sets out
to explain DMN function. Indeed, it is known that, the more the external world is predictable,
the more mental activity becomes detached from the actual sensorium
\citep{antrobus1966studies, pope1978regulation}.
Conversely, the more the currently executed task is unknown and unpracticed,
the lower the DMN activity, i.e the tendency for stimulus-independent thoughts
\citep{filler1973daydreaming, teasdale1995stimulus}.

\subsection{Value functions and policies}
The behavior of the agent is governed by some kind of \textit{policy}, which maps states to
sampling probabilities over candidate actions to perform from this state. Starting a time $t=0$,
a policy $\pi$ generates a trajectory as follows
\begin{eqnarray*}
  \begin{split}
    S_1 &\sim p(.|S_0,A_0), \text{ collect rewards }R_0 = r(S_0, A_0)\\
    A_1 &\sim \pi(.|S_1)\\
    S_2 &\sim p(.|S_1,A_1), \text{ collect rewards }R_1 = r(S_1, A_1)\\
    \ldots\\
    A_{t} &\sim \pi(.|S_{t})\\
    S_{t+1} &\sim p(.|S_{t},A_{t}), \text{ collect rewards }R_{t} = r(S_{t}, A_{t})\\
    \ldots
  \end{split}
\end{eqnarray*}
Since an action taken now might have repercussions long into the feature, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
cumulative reward estimate which takes into account the future. A standard for
modeling this accumulation is the so-called
time-discounted \textit{cumulative reward} function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t.
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic).}  measures the cumulative reward of
following a
policy $\pi$.
\begin{mdframed}
  \textbf{Where is ``value'' buffered in the brain ?}
  The amygdala is known to be involved in significance evaluation) and
  the nucleus accumbens is involved in reward evaluation.
  \end{mdframed}
The  goal of RL is then adapt this policy so as to  maximize the cumulative rewards.
In the definition of $G^\pi$ above, the constant $\gamma \in [0, 1]$ is the \textit{discount factor}.
Taking $\gamma = 0$   corresponds to a perfectly myopic agent who is solely concerned about their
immediate rewards and has no horizon for the future, which is not compatible with long-term goal planning
as implemented in the DMN. For there to be any learning at all, it is thus necessary that $\gamma > 0$.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
Thus the agent is positively biased towards considering rewards which are will
come soon. The reward importance hence decays exponentially
with distance in the future,
given that there is more uncertainty in the farsighted future.
The reward discount factor $\gamma$ can thus be seen as calibrating
the delay discounting behavior of the intelligent agent.
It is also important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
For instance, a deterministic strategy in playing rock-paper-scissors can
be easily exploited and a uniformly random choice is optimal.

@@
